---
title: "STSCI4850 HW2"
output: html_document
date: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

# Load library
```{r,ECHO = FALSE, include=TRUE}
library(tidyverse) # Load the tidyverse
```

# Load the data
```{r}
organ_data <- read_csv("global-organ-donation_2018.csv")
```


# Basic review

```{r}
# number of observation
nrow(organ_data)

# what does each row mean? OR, what is the unit of inference?
# Each row represents a country's reported metrics for organ donation for a given year 

# How many variables?
ncol(organ_data)
# list the variable names
colnames(organ_data)
# How many years
unique(organ_data$REPORTYEAR)|>
  length()
# countries
unique(organ_data$COUNTRY)|>
  length()
# Is it a balanced panel? 
organ_data|> count(COUNTRY)|>
  arrange(desc(n))
# No this is not a balanced panel as not every country has the same number of observations
```

# keep variables REGION through "Actual DCD"

```{r}
#meaning of variables
# REGION: The region of the reported data
# COUNTRY: The country of the reported data
# REPORTYEAR: The year of the data data
# POPULATION: The reported data's country population in millions
# TOTAL Actual DD: Total number of deceased donors
# Actual DBD: Actual Donation after a donor's brain death
# Actual DCD: Actual Donation after donor's circulatory death
# Total Utilized DD: Donors who donated at least one organ that was transplanted into a recipient
# Utilized DBD: Brain death donors whose organs were used for transplant
# Utilized DCD: Circulatory death donors whose organs were used for transplant

#units of measurement
# Population is in millions 
# Donors are regular raw counts

#What are the observational units?
# The observational units are country year units

#messy data

#Invalid or inconsistent values
na_count <- sapply(organ_data, function(x) sum(is.na(x)))
na_count

#Improperly formatted missing values
#Nonstandard data format
#Messy column names
#Improper variable types
#Incomplete data


# change names to lowercase and meaningful names
library(janitor)

organ_data_clean <- organ_data|>
  clean_names() |>
  rename(
    report_year = reportyear,
    total_deceased_donors = total_actual_dd,
    brain_dead_donors = actual_dbd,
    circulatory_death_donors = actual_dcd,
    utilized_brain_dead_donors = utilized_dbd,
    utilized_circulatory_death_donors = utilized_dcd,
    deceased_kidney_donors = dd_kidney_tx,
    living_kidney_donors = ld_kidney_tx,
    total_kidney_donors = total_kidney_tx,
    deceased_liver_donors = dd_liver_tx,
    living_liver_donors = ld_liver_tx,
    domino_liver_donors = domino_liver_tx,
    total_liver_donors = total_liver_tx,
    deceased_lung_donors = dd_lung_tx,
    living_lung_donors = ld_lung_tx,
    total_lung_donors = total_lung_tx,
    total_heart_donors = total_heart,
    total_pancreas_donors = pancreas_tx,
    total_kidney_pancreas_donors = kidney_pancreas_tx,
    total_small_bowel_donors = small_bowel_tx
  )
# why is population in decimals?
# Population is in decimals to represent the population in millions without having to write out so many digits

# Is total deceased equal to sum of deceased by brain and cardiac death? Comment 
organ_data_clean |> 
  mutate(
    total_brain_circulatory = brain_dead_donors + circulatory_death_donors,
    difference = total_deceased_donors - total_brain_circulatory
  ) |>
  filter(!is.na(total_deceased_donors)) |> 
  select(country, report_year, total_deceased_donors, total_brain_circulatory, difference) |>
  head(10)

# There is a discrepancy between the noted total and the of brain and circulatory calculated sum. This exists because a lot of the countries report the total number of deceased donors but are unable to or don't note the sub categories of brain dead or circulatory death donors. Some countries' healthcare systems might have different definitions for death that don't fit the classifications of DBD or DCD. The column that notes the total deceased donors would be the most reliable metric.

```

# Reproduce fig 4.2

```{r}

organ_data_clean |>
  filter(!is.na(total_deceased_donors)) |>
  group_by(report_year) |>
  summarise(
    count = n()
  ) |>
  ggplot(aes( x = report_year, y = count)) +
  geom_col() +
  labs(
    x = "Year",
    y = "Nonmissing TOTAL Actual DD values",
    title = "Reproduction of Figure 4.2"
  ) 
```

# Reproduce fig 4.3

```{r}

organ_data_clean |>
  group_by(report_year) |>
  summarise(total = n()) |>
  ggplot(aes( x = report_year, y = total)) +
  geom_col() +
  labs(
    x = "Year",
    y = "Number of rows",
    title = "Reproduction of Figure 4.3"
  ) 

```


# Reproduce fig 5.2 

```{r}
source("imputeFeature.R")

# Complete the data so that balanced panel: 

organ_balanced_data <- organ_data_clean |>
  complete(country, report_year)

# Use "average" for imputation

organ_data_imputed <- organ_balanced_data |>
  mutate(total_deceased_donor_imputed = imputeFeature(
    .data = pick(everything()),
    .feature = total_deceased_donors, 
    .group = country, 
    .impute_method = "average"
  ),
  population_imputed = imputeFeature(
    .data = pick(everything()),
    .feature = total_deceased_donors, 
    .group = country, 
    .impute_method = "average"
  )) 

organ_data_imputed |>
  group_by(report_year) |>
  summarise(global_total = sum(total_deceased_donor_imputed, na.rm = TRUE)) |>
  ggplot(aes(x = report_year, y = global_total)) +
  geom_line() +
  labs(
    x = "Year",
    y = "(imputed) Global number of organ donors",
    title = "Reproduction of Figure 5.2"
  )
```


# Reproduce fig 5.7 

```{r}
organ_data_2017 <- organ_data_imputed |>
  filter(report_year == 2017) |>
  mutate(total_deceased_donors_per_mil = total_deceased_donor_imputed / (population + 1) * 1000000000,
         total_deceased_donors_per_mil = replace_na(total_deceased_donors_per_mil, 0))

organ_data_2017 |>
  ggplot(aes(x = total_deceased_donors_per_mil)) +
  geom_histogram() +
  labs(
    x = "(Imputed) organ donors per million in 2017",
    y = "Number of countries",
    title = "Reproduction of Figure 5.7"
  )

```


# Investigate imputation

```{r}
# Create a figure comparable to fig 5.2 using raw data i.e., without imputation. Note this means you have to remove missing values


# Why does donor count drops sharply after 2014 in raw data? 


# with imputation data
# print the total (across all countries) raw and imputed deceased total count for China
# print the raw and imputed deceased total count for Peru
organ_data_imputed |> 
  filter(country == "Peru") |> 
  summarise(
    raw_total_china = sum(total_deceased_donors, na.rm = TRUE), 
    imputed_total_china = sum(total_deceased_donor_imputed)
  )

# print the raw and imputed deceased total count for China

organ_data_imputed |> 
  filter(country == "China") |> 
  summarise(
    raw_total_china = sum(total_deceased_donors, na.rm = TRUE), 
    imputed_total_china = sum(total_deceased_donor_imputed)
  )

# comment on the imputation method used - would you suggest any changes

#The donor count sharply drops in raw data counts after 2014 is probably because of a lag or slow down in reporting. Data may not have been verified or submitted to the database yet. The imputation method fills the gaps using an average of the country's prior stats. This smooths the trend line and removes cliffs such as the one that occured in the raw data. This is a good approach for missing data for missing years, but it fails to completely acknowledge with edge cases at the start or end of the data timeline that has no future data to average with. Including regional growth rates or regression models that show development trajectories for different countries instead of just using the last known statistic could improve accuracy.

organ_data_clean |>
  group_by(report_year) |>
  summarise(raw_global_total = sum(total_deceased_donors, na.rm = TRUE)) |>
  ggplot(aes(x = report_year, y = raw_global_total)) +
  geom_line() +
  labs(
    x = "Year",
    y = "Global number of organ donors (Raw)",
    title = "Raw Data Worldwide Organ Donor Trends"
  )

```

# Reproduce Table 1
# Fig 5.2 or Table 5.1 Which is a better representation? comment

```{r}

  # # note that we use `population_imputed + 1` in the denominator because there 
  # # are some countries with a reported population of 0.
  # mutate(total_deceased_donors_per_mil = total_deceased_donors / (population_imputed + 1) * 1000000,
  #        total_deceased_donors_imputed_per_mil = total_deceased_donors_imputed / (population_imputed + 1) * 1000000) 


```
The graph reproducing figure 5.2 is a better representation as it gives a visual comparison between the raw lagged reporting data and how imputation is able to piece together the global trend. A table can give numbers that seem more precise, but the line graph more clearly shows the data, showing the decline after 2014 (which is artificial) more obvious see. Ultimately, the visualization highlights a constant and steady increase organ donation globally that could easily be overlooked in a tabular format.

# Reproduce Fig 5.6 and comment
```{r}

top_15 <- organ_data_imputed |>
  ungroup() |>
  filter(report_year == 2017) |>
  mutate(rate = total_deceased_donor_imputed / (population_imputed + 1)) |>
  slice_max(rate, n = 15) |>
  pull(country)

top_15_data <- organ_data_imputed |>
  filter(country %in% top_15) |>
  mutate(rate = total_deceased_donor_imputed / (population_imputed + 1))

ggplot(top_15_data, aes(x = report_year, y = rate)) +
  geom_line() +
  facet_wrap(~country) +
  ggtitle("Number of organ donations per million from 2000 to 2017",
          subtitle = "For the top 20 countries in 2017")


```


# Which countries showed the highest growth rate in donor rates? Comment

```{r}

```
