---
title: "NHANES Hypertension"
output: pdf
date: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## replicate the sample size in https://www.cdc.gov/nchs/data/databriefs/db519.pdf


# Load library
```{r,ECHO = FALSE, include=TRUE}
rm(list=ls())

library(tidyverse) # Load the tidyverse
library(survey)

```

#Replicate Figure 1, Table 1
Sample size is unweighted. Estimates are weighted using fasting sample weights. Estimates for diagnosed diabetes are based on responses to the
survey question, “Other than during pregnancy, have you ever been told by a doctor or health professional that you have diabetes or sugar diabetes?” Estimates
for undiagnosed diabetes are based on an 8- to 24-hour fasting plasma glucose greater than or equal to 126 mg/dL or hemoglobin A1c greater than or equal to
6.5% in a participant who reported never receiving a diabetes diagnosis from a healthcare provider. Age-adjusted estimates for adults age 20 and older were age
adjusted by the direct method to the U.S. Census 2000 population using age groups 20–39, 40–59, and 60 and older

all data from: https://wwwn.cdc.gov/nchs/nhanes/default.aspx
2021-2023

# Load the data
```{r}
demo <- haven::read_xpt("DEMO_L.xpt")
bpxo <- haven::read_xpt("BPXO_L.xpt")
bpq  <- haven::read_xpt("BPQ_L.xpt")
 
df1 <- demo %>%
  select(SEQN, RIAGENDR, RIDAGEYR, RIDEXPRG, RIDRETH3,
         WTMEC2YR,   # ← MEC exam weight for the 2021-2023 cycle
         SDMVSTRA, SDMVPSU) %>%
  rename(gender = RIAGENDR,
         age    = RIDAGEYR,
         preg   = RIDEXPRG,
         race   = RIDRETH3,
         weight = WTMEC2YR)  # renamed for convenience
 
df2 <- bpxo %>%
  select(SEQN,
         BPXOSY1, BPXOSY2, BPXOSY3,
         BPXODI1, BPXODI2, BPXODI3)
 
df3 <- bpq %>%
  select(SEQN,
         BPQ020,   # Ever told you had high blood pressure
         BPQ150)   # Currently taking high blood pressure medication
         # NOTE: BPQ150 in the L (2021-2023) cycle corresponds to "Are you now
         # taking any medication prescribed by a doctor for your high BP?"
         # In earlier cycles this was BPQ040A — verify against your codebook.
 
tmp <- df1 %>%
  merge(df2, by = "SEQN") %>%
  merge(df3, by = "SEQN")


```
We load the data and merge each important variable with their ...



```{r}
# ── Before cleaning, understand what you have ─────────────────────────────
df <- tmp %>%
 
  # -- Restrict to adults 18+ ------------------------------------------------
  filter(age >= 18) %>%
 
  # -- Keep only MEC-examined participants ------------------------------------
  #    weight == 0 means they did not attend the mobile exam center
  filter(weight > 0) %>%
 
  # -- Exclude pregnant women (per CDC methods) --------------------------------
  filter(is.na(preg) | preg != 1) %>%

  # -- Recode refused/don't know to NA ----------------------------------------
  #    NHANES codes: 7 = Refused, 9 = Don't know
  mutate(
    BPQ020 = if_else(BPQ020 %in% c(7, 9), NA_real_, BPQ020),
    BPQ150 = if_else(BPQ150 %in% c(7, 9), NA_real_, BPQ150)
  ) %>%
 
  # -- Average SBP and DBP across readings 2 and 3 ----------------------------
  #    Reading 1 is a practice read per NHANES protocol and is excluded.
  #    CDC uses "average of up to three" — they include reading 1 only when
  #    readings 2 and 3 are both missing. Replicate with rowMeans(na.rm=TRUE)
  #    applied only to readings 2 and 3 to match the standard approach.
  mutate(
    sbp = rowMeans(cbind(BPXOSY2, BPXOSY3), na.rm = TRUE),
    dbp = rowMeans(cbind(BPXODI2, BPXODI3), na.rm = TRUE)
  ) %>%
   filter(!is.nan(sbp) & !is.nan(dbp)) %>%
  # -- On medication flag (used to define hypertension) -----------------------
  #    BPQ150: 1 = Yes, 2 = No, NA = not asked (never diagnosed → not on meds)
  mutate(
    on_meds = case_when(
      BPQ150 == 1           ~ 1,
      BPQ150 == 2           ~ 0,
      is.na(BPQ150)         ~ 0,   # question not asked → not on medication
      TRUE                  ~ 0
    )
  ) %>%
 
  # -- OUTCOME 1: Hypertension ------------------------------------------------
  #    SBP >= 130 OR DBP >= 80 OR currently taking BP medication
  mutate(
    hypertension = if_else(sbp >= 130 | dbp >= 80 | on_meds == 1, 1, 0)
  ) %>%
 
  # -- OUTCOME 2: Awareness ---------------------------------------------------
  #    DENOMINATOR: adults WITH hypertension
  #    DEFINITION: ever told by a doctor (BPQ020 == 1)
  mutate(
    aware = case_when(
      hypertension == 1 & BPQ020 == 1 ~ 1,
      hypertension == 1 & BPQ020 == 2 ~ 0,
      TRUE ~ NA_real_          # not hypertensive → exclude
    )
  ) %>%
 
  # -- OUTCOME 3: Treatment ---------------------------------------------------
  #    DENOMINATOR: adults WITH hypertension   ← KEY FIX vs. your original code
  #    DEFINITION: currently taking prescribed BP medication
  #    NOTE: Anyone on_meds == 1 is already hypertensive by definition above,
  #    so treated == 1 is a subset of hypertension == 1.
  mutate(
    treated = case_when(
      hypertension == 1 & on_meds == 1 ~ 1,
      hypertension == 1 & on_meds == 0 ~ 0,
      TRUE ~ NA_real_          # not hypertensive → exclude
    )
  ) %>%
 
  # -- OUTCOME 4: Control -----------------------------------------------------
  #    DENOMINATOR: adults WITH hypertension   ← KEY FIX vs. your original code
  #    (CDC Table 4 sample sizes match Table 1 hypertensive counts, NOT
  #     treatment counts. Control is NOT restricted to treated individuals.)
  #    DEFINITION: SBP < 130 AND DBP < 80
  mutate(
    controlled = case_when(
      hypertension == 1 & (sbp < 130 & dbp < 80)   ~ 1,
      hypertension == 1 & (sbp >= 130 | dbp >= 80) ~ 0,
      TRUE ~ NA_real_          # not hypertensive → exclude
    )
  ) %>%
 
  # -- Age groups -------------------------------------------------------------
  mutate(
    agegroup = case_when(
      age >= 18 & age <= 39 ~ "18-39",
      age >= 40 & age <= 59 ~ "40-59",
      age >= 60             ~ "60+",
      TRUE ~ NA_character_
    ),
    agegroup = factor(agegroup, levels = c("18-39", "40-59", "60+"))
  ) %>%
 
  # -- Sex label --------------------------------------------------------------
  mutate(
    sex = factor(if_else(gender == 1, "Male", "Female"),
                 levels = c("Male", "Female"))
  ) %>%
 
  # -- Race/ethnicity ---------------------------------------------------------
  mutate(
    race_eth = case_when(
      race == 1 ~ "Mexican American",
      race == 2 ~ "Other Hispanic",
      race == 3 ~ "Non-Hispanic White",
      race == 4 ~ "Non-Hispanic Black",
      race == 6 ~ "Non-Hispanic Asian",
      race == 7 ~ "Other/Multiracial",
      TRUE ~ NA_character_
    ),
    race_eth = factor(race_eth)
  )
 
# -- Sanity check: should be a descending funnel ------------------------------
df %>%
  summarise(
    n_total        = n(),
    n_hypertension = sum(hypertension == 1,   na.rm = TRUE),
    n_aware        = sum(aware == 1,           na.rm = TRUE),
    n_treated      = sum(treated == 1,         na.rm = TRUE),
    n_controlled   = sum(controlled == 1,      na.rm = TRUE)
  )
# Expected (unweighted): ~6084 total, ~3242 hypertensive
# Weighted estimates should approximate: 47.7% / 59.2% / 51.2% / 20.7%
```

```{r}

  # -- OUTCOME 1: Hypertension ------------------------------------------------
  #    SBP >= 130 OR DBP >= 80 OR currently taking BP medication
  df %>%
  
  mutate(
    hypertension = if_else(sbp >= 130 | dbp >= 80 | on_meds == 1, 1, 0)
  ) %>%
 
  # -- OUTCOME 2: Awareness ---------------------------------------------------
  #    DENOMINATOR: adults WITH hypertension
  #    DEFINITION: ever told by a doctor (BPQ020 == 1)
  mutate(
    aware = case_when(
      hypertension == 1 & BPQ020 == 1 ~ 1,
      hypertension == 1 & BPQ020 == 2 ~ 0,
      TRUE ~ NA_real_          # not hypertensive → exclude
    )
  ) %>%
 
  # -- OUTCOME 3: Treatment ---------------------------------------------------
  #    DENOMINATOR: adults WITH hypertension
  #    DEFINITION: currently taking prescribed BP medication
  #    NOTE: Anyone on_meds == 1 is already hypertensive by definition above,
  #    so treated == 1 is a subset of hypertension == 1.
  mutate(
    treated = case_when(
      hypertension == 1 & on_meds == 1 ~ 1,
      hypertension == 1 & on_meds == 0 ~ 0,
      TRUE ~ NA_real_          # not hypertensive → exclude
    )
  ) %>%
 
  # -- OUTCOME 4: Control -----------------------------------------------------
  #    DENOMINATOR: adults WITH hypertension
  #    (CDC Table 4 sample sizes match Table 1 hypertensive counts, NOT
  #     treatment counts. Control is NOT restricted to treated individuals.)
  #    DEFINITION: SBP < 130 AND DBP < 80
  mutate(
    controlled = case_when(
      hypertension == 1 & (sbp < 130 & dbp < 80)   ~ 1,
      hypertension == 1 & (sbp >= 130 | dbp >= 80) ~ 0,
      TRUE ~ NA_real_          # not hypertensive → exclude
    )
  ) %>%
 
  # -- Age groups -------------------------------------------------------------
  mutate(
    agegroup = case_when(
      age >= 18 & age <= 39 ~ "18-39",
      age >= 40 & age <= 59 ~ "40-59",
      age >= 60             ~ "60+",
      TRUE ~ NA_character_
    ),
    agegroup = factor(agegroup, levels = c("18-39", "40-59", "60+"))
  ) %>%
 
  # -- Sex label --------------------------------------------------------------
  mutate(
    sex = factor(if_else(gender == 1, "Male", "Female"),
                 levels = c("Male", "Female"))
  ) %>%
 
  # -- Race/ethnicity ---------------------------------------------------------
  mutate(
    race_eth = case_when(
      race == 1 ~ "Mexican American",
      race == 2 ~ "Other Hispanic",
      race == 3 ~ "Non-Hispanic White",
      race == 4 ~ "Non-Hispanic Black",
      race == 6 ~ "Non-Hispanic Asian",
      race == 7 ~ "Other/Multiracial",
      TRUE ~ NA_character_
    ),
    race_eth = factor(race_eth)
  )
 
```
BPQ020 - Ever told you had high blood pressure
BPQ030 - Told had high blood pressure - 2+ times
BPQ150 - Taking high blood pressure medication
BPQ080 - Doctor told you - high cholesterol level
BPQ101D - Taking meds to lower blood cholesterol?



```{r}

# -- Sanity check: should be a descending funnel ------------------------------
df %>%
  summarise(
    n_total        = n(),
    n_hypertension = sum(hypertension == 1,   na.rm = TRUE),
    n_aware        = sum(aware == 1,           na.rm = TRUE),
    n_treated      = sum(treated == 1,         na.rm = TRUE),
    n_controlled   = sum(controlled == 1,      na.rm = TRUE)
  )
# Expected (unweighted): ~6084 total, ~3242 hypertensive
# Weighted estimates should approximate: 47.7% / 59.2% / 51.2% / 20.7%



# Instead of this (manual):
est  <- svymean(~hypertension, svy, na.rm = TRUE)
pct  <- coef(est) * 100
se   <- SE(est)   * 100
ci_l <- pct - 1.96 * se
ci_u <- pct + 1.96 * se

# You can do this (automatic):
result <- svyby(~hypertension, ~sex, svy, svymean, na.rm = TRUE, vartype = c("se", "ci"))


```